Landcover Data Limits Our Understanding of Earth System Processes

### 2015-10-22 paper/main.tex
22/10/2015 revision. Revisions through results. See changelog.

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,sort,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
\usepackage{placeins}
\usepackage{float}
%\usepackage{subfigure}
\usepackage{wrapfig}
%\setlength\belowcaptionskip{-2cm}
%\setlength{\textfloatsep}{5pt plus 1.0pt minus 2.0pt}

@@

\title{A Full Accounting of Landcover Map Error and Bias and Their Impacts on Assessments of Global Change}
%\title{Our Understanding of Global Change is Built on a Shaky Foundation}

@@

\dropcap{T}he nature and distribution of landcover provides significant insight into econonomic processes \cite{lambin_modelling_1997} because human endeavors are so closely tied to how we transform land, whether it be the felling of ancient forests for farmland or converting that farmland into office parks. The vastness of our alteration of Earth’s landscapes suggests that landcover is a prime mediator of many environmental and social processes that drive or are affected by global change \cite{lambin_modelling_1997}, such as agricultural production and food security \cite{lark_cropland_2015,wright_recent_2013, licker_mind_2010}, carbon cycling \cite{asner_high-resolution_2010,gaveau_major_2014}, biodiversity loss \cite{newbold_global_2015,luoto_predicting_2004}, and changes in human demography \cite{linard_assessing_2010}. Like any view into nature, resolution and fidelity at fine scales are the keys to unlocking more granular and mechanistic insights into these processes \cite{see_improved_2015}. It is therefore unsurprising to see the explosive growth in private sector initiatives to develop new Earth observing capabilities, which range from small hobbyist drones\footnote{e.g. 3DRobotics, DJIA} to satellite arrays\footnote{Planet Labs, Skybox}, in order to add value to industries such as agriculture, mining, and construction. This rapid growth in fine-scale landcover mapping capability is creating new opportunities to develop actionable information for traditionally public-sector concerns, such as agricultural development\footnote{USAID's Feed the Future}, drought and flood adaptation\footnote{Global Index Insurance Facility, www.indexinsuranceforum.org}, and carbon cycle management\footnote{United Nations REDD+, www.un-redd.org/aboutredd}. But while the demand for more nuanced, landcover-based insights is growing, there is only now the opportunity to use finer-scaled imagery to comprehensively interrogate the accuracy and biases in the landcover products that have become ubiquitous in global change research.

Global landcover data can only practically be derived from satellite imaging, but in many regions the cover types of interest are smaller, on average, than the sensor resolution, or spectrally indistinct from other neighboring covers, and these factors propagate classification errors \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. The result is that landcover maps are generally inaccurate at finer scales and disagree substantially with one another, particularly in those parts of the world undergoing the most rapid land use changes \cite{estes_projected_2013,fritz_comparison_2010,fritz_cropland_2011}.

Errors in landcover products are widely-acknowledged \cite{fritz_comparison_2010, fritz_cropland_2011, see_improved_2015, fritz_mapping_2015,verburg_challenges_2011}, and there are a variety of efforts underway to improve landcover maps, particularly for agriculture \cite{fritz_geo-wiki:_2012,estes_diylandcover:_2015}. What is less known is the degree to which these errors bias analyses derived from the distributional and areal information in landcover. Errors are hard to quantify because spatially extensive reference data are not available for most regions of the world--particularly over Africa and other developing regions. Error assessments therefore typically rely on a small number of ground truth points for a bottom-up assessment or aggregated survey data for a top-down sanity check. For this reason, we have a better understanding of discrepancies between landcover datasets in relation to country-level statistics \cite{fritz_comparison_2010,fritz_cropland_2011,kaptue_tchuente_comparison_2011}, which offers little direction for how to arrive at a true number.

Being unable to fully quantify the errors in landcover maps of course makes it difficult, if not impossible, to quantify their impact on downstream analyses. There has been some work examining how such error influences climate simulations \cite{ge_impacts_2007}, agricultural land use patterns \cite{schmit_limitations_2006}, and carbon flux \cite{quaife_impact_2008} and human population estimates \cite{linard_assessing_2010}, but these either use simulated landcover errors \cite{ge_impacts_2007} or compare relevant differences in estimates between different satellite-derived landcover maps \cite{linard_assessing_2010, quaife_impact_2008}. One exception is a Belgian study  \cite{schmit_limitations_2006} that used ground-collected farm parcel data to assess how landcover map errors bias measurements of agricultural land use patterns, but the study extent was fairly small and the validation data were discontiguous.

Just as a building needs a solid foundation, global change science needs to be based on sound landcover data. There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need here, using a unique, high accuracy agricultural landcover map for the entire country of South Africa to conduct a spatially comprehensive, bottom-up quantification of error in several latest generation landcover maps that are widely used in global change studies. We use these errors to assess the biases and accuracies of i) these landcover data, ii) how landcover properties influence map errors, iii) how bias and accuracy change with aggregation scale, with the specific goal of determining ``safe'' scales for drawing area-based inferences, and iv) how landcover error propagates through ``downstream'' analyses that represent several major global change research areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

\vspace{-0.5 cm}
\section{Study area and landcover data}
South Africa comprises nearly 6\% of sub-Saharan Africa's (SSA) landmass, and has a large, diverse agricultural sector, ranging from large commercial operations to smallholder farms \cite{hardy_rainfed_2011,estes_using_2014}. This diversity suggests that the country's agricultural landcover spans the range of types that are found throughout the rest of SSA.

We compared the reference map with similar maps derived from four existing landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite[e.g.][]{fry_completion_2009}, MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013, shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and represents the current state-of-the-art in landcover mapping. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates, resulting in 4 maps (hereafter simply the ``cropland maps'') to compare to our reference map.

\vspace{-0.5 cm}
\section{Quantifying Error and Bias}
We first calculated errors in pixel-wise estimated cropland percentages by differencing the reference map with each cropland map at each of the five aggregation scales (1-100 km). We used these errors to calculate map bias (the mean pixel-wise error) and accuracy (the mean absolute error, and how these vary with scale. Next, we assessed the degree to which the average cropland cover in agricultural landscapes, a descriptor of landcover pattern, impacts map accuracy.

To investigate how landcover map error can impact downstream global change research, we quantified the biases and accuracies in five landcover-based analyses built upon our reference and cropland maps. The first three represented ``first-order'' analyses, in which a variable of interest is mapped onto a landcover type(s) using a simple empirical relationship. The first of these was the widely used International Panel on Climate Change's Tier-1 approach for mapping vegetative carbon stocks, as developed by \cite{ruesch_new_2008}. The second was maize yield maps derived by disaggregating district-scale agricultural census data for both maize yield and harvested area \cite[following][]{monfreda_farming_2008,ramankutty_farming_2008}, from which we calculated the third map, gridded maize production estimates. Maps based on these analyses underpin many assessments of crop productivity and production \cite[e.g.][]{foley_solutions_2011,licker_mind_2010}.

The other two analyses can be considered second-order, wherein a process model draws on the cover types' values to calculate an output value. For the first of these, we used the Variable Infiltration Capacity \cite{liang_simple_1994} land surface hydrology model to calculate monthly evapotranspiration, using the reference and cropland maps to adjust landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these map errors impact the land allocation process of an agent-based food security model \cite{chen_dependency_2013}.

\vspace{-0.5 cm}
\section{Landcover maps}
\subsection{Bias and accuracy}

We created the 1 km reference map after removing all field types classified as communal/smallholder agriculture (individual fields in this category were not delineated, thus they were removed to prevent potential cropland area overestimates, SI) or permanent tree-crops (SI), and calculated the total cropland extent in the remaining area (1,081,000 km$^2$, or 90\% of South Africa). The 2011 reference map showed a cropland area of 104,304 km$^2$, which the SA-LC and GeoWiki maps overestimated by 31 and 10\%, respectively, and GlobCover and MODIS underestimated by 18 and 23\%.

We then aggregated the reference and each cropland map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from each reference (2007 and 2011) map at each scale of aggregation to assess error patterns (Fig. 1A). Negative pixels here represent overestimation error, while positive values indicate underestimates.

%\vspace{-0.5 cm}
\begin{figure*}[ht]
%\begin{wrapfigure}{c}{1\textwidth}
\centerline{\includegraphics[width=0.99\textwidth]{figures/figure1.pdf}}
\caption{(A) Errors in the percent cropland estimates resulting from each of the four cropland maps relative to the reference map at different scale of pixel aggegation. Rows indicate the landcover map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas with no data where communal farmlands or plantation forests were removed from analysis. (B) The bias (mean error) and accuracy (mean absolute error [MAE]) of each cropland map at each scale of aggregation. Bias estimates (represented by symbols) fall within the semi-transparent bars, mean absolute errors in the solid bars, with bar colors coded to specific cropland maps. The symbols indicate the method of calculating each metric, either by averaging across the entire country, within the union of  agricultural areas (cropland $>$0.05\%) for each reference-cropland map combination, or as the average of the averages calculated per 5\% bin of cropland cover, a measure of error that is independent of cropland density.}\label{afoto}
\end{figure*}
%\end{wrapfigure}

@@

\linenumbers
The most pronounced errors were in the MODIS and GlobCover maps, which both underestimated cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1A, and the dominant production region), and overestimated along the eastern to northern margins (red areas in Fig. 1A). All maps have less than +/-5\% bias (crosses in semi-transparent bars, Fig. 1B), at all aggregation scales, when errors are averaged across the entire country, due to the extensive non-agricultural areas in the country that were effectively discerned by all maps. The whole country mean absolute error (the measure of map accuracy) were a few percent higher values in most cases, but still $<$10\%, except for GlobCover ($\sim$11\%) at 1 km resolution. Biases increase only slightly when errors are averaged within the combined agricultural regions (areas with greater than 0.05\% cropland, indicated by circles in transparent bars in Fig. 1B) of the reference and each cropland map, but accuracy decreases substantially, with mean absolute errors increasing to 10\% for SA-LC and around 20\% for MODIS, GlobCover, and GeoWiki at 1 km resolution (circles in solid bars in Fig. 1B), and remain between 5-10\% for all three datasets even at 25 km of aggregation.

As a third measure, we calculated bias and mean absolute error by averaging their mean per-ventile values (i.e. the average bias or accuracy value for areas have 0-5\% cropland, 5-10\% cropland, etc.). These provide density independent measures of map performance because they remove the dominance any particular level of cover has on map accuracy, which in this case was the 0-5\% cover class where map performance was best (SI). Calculated this way, bias and accuracy were substantially worse. The average bias jumped to 21\% for MODIS and 34\% for GlobCover at 1 km resolution (Fig. 1B), meaning that each map had a strong tendency to underestimate cropland. This decreased with pixel aggregation, falling to 8\% at 50 km for MODIS, but remaining as high as 14\% at 100 km for GlobCover. SA-LC bias was smaller in magnitude, but tended towards cropland overestimates of 5-10\% at all aggregation scales. GeoWiki alone was relatively unbiased, but showed substantial inaccuracy with a density-independent mean absolute error of 23\%, which dropped below 10\% only above 25 km of aggregation. GlobCover was by far the most inaccurate, having $>$35\% mean absolute error at 1 km and 17\% at 100 km, following by MODIS (over 30\% at 1 km and 11\% at 100 km). SA-LC was the most accurate of all four maps with the density-independent measure, ranging from just above 10\% mean absolute error at 1 km to 5\% at 100 km.

\subsection{Landscape characteristics and error}
The density-independent measures indicate that error and bias increase with cropland density, which suggests that map error varies as a function of the typical abundance of cover in the landscape being imaged. To investigate the relationship between cover density and map accuracy, we calculated the mean density for agricultural pixels ($>$0.05\% cropland) in the 1 km reference map within the boundaries of 354 magisterial districts (South Africa's finest administrative unit, averaging 3,445 km$^2$; SI Figure S3), providing a landscape-scale measure representing typical cropland density. We then extracted the cropland map errors for these same pixels, and calculated their district-wise mean absolute errors.

A generalized additive model fit to district-level mean absolute error (log-transformed) shows that error peaks at 50-60\% cropland cover for all but the GlobCover map (which continued to increase with cropland cover), and is lowest when the landscape is dominated either by cropland or other cover types (Fig. 2). In other words, accuracy is generally lowest when cropland cover is mixed evenly with other cover types. GlobCover's accuracy continued to decrease with cropland density because the dominant cover class contributing to the percentage cropland estimate was a mixed class defined as 50-70\% crops mingled with other vegetation, thus the maximum percentage was constrained by this mixture range.

%\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between map accuracy (the mean absolute error) in cropland maps and the actual cropland cover within agricultural landscapes (reference map pixels having $>$0.5\% cropland), here defined by the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different cropland maps, with the solid line indicating predicted absolute bias, and the lighter shading the standard error of the coefficients.}
\label{afoto}
\end{figure}

\section{First-order analyses}
\subsection{Tier-1 carbon estimates}
Using the methods provided by \cite{ruesch_new_2008}, we calculated average carbon densities for African forests, secondary forests, shrublands, croplands, grasslands, and sparse habitats (semi-arid grasslands and low shrublands), and assigned cropland carbon values to map cells in proportion to their cropland cover. For the non-cropland proportions, we assigned the carbon value from each of the other types, creating five different carbon maps for each landcover map at each aggregation scale (Fig. S4), which allowed us to test how carbon estimates vary as a function of i) cropland map bias and ii) the characteristics of adjacent cover types.

The difference between total carbon stocks for the country made using any of the cropland maps were within +/-3\% of those based on the reference map, regardless of which cover type was adjacent to cropland (Table S1), because the large of non-cropland in the country ($\sim$50-70\%, Fig. 1) dilutes any map errors. Comparing total stocks between maps for just the agricultural area (30-50\% of the country) reveals much greater differences (Table S1). SA-LC overestimated carbon stocks by just 2\% when the adjacent cover type was forest, and up to 15\% when it was sparse cover. MODIS ranged from negligible differences in denser carbon classes (forest, secondary forest, and shrublands) to 8-13\% underestimates for the grassland and sparse classes. GeoWiki underestimated for all types, from $<$1\% for sparse cover to 8\% for forest. GlobCover grossly overestimated total carbon stocks for agricultural areas, varying from 64\% for sparse lands to 162\% for forest. The magnitude of this bias was due to false positives--GlobCover identified cropland in nearly 50\% of pixels, compared to 30\% for the other three cropland maps.

The spatial patterns of errors in carbon estimates (Fig. S4) reflect those of the cropland maps (Fig. 1). Where cropland was underestimated and the surrounding cover type was of higher carbon density than cropland, carbon density was overestimated. For lower density cover (grassland and sparse vegetation), carbon stocks were underestimated, but by small magnitudes. These tendencies were reflected in each map's biases, as calculated over the cropped areas of the country as jointly defined by the reference and each cropland map (Fig. 3). For example, MODIS and GlobCover bias was $\sim$-50\% (overestimation) at 1 km resolution when forest was the cropland-adjacent cover (stars in semi-transparent green and gold bars, Fig. 3; Table S2). For sparse vegetation (open circles in Fig. 3), MODIS bias was 3\% at all scales, whereas maps that overestimated cropland (e.g. SA-LC, semi-transparent red) overestimated carbon density for this cover type, because cropland has a higher carbon density \cite{ruesch_new_2008}. Overall, GeoWiki had the lowest bias, for all cover types and all resolutions. Its worst bias was a tendency to overestimate by 12\% at 1 km when forest was adjacent, but at coarser scales this bias reduced to just a few percent (Fig. 3, Table S2). All maps' biases are within +/-10\% bias after aggregation to 25 km.

%%\vspace{-0.9 cm}
%\begin{figure}[!ht]
%\centerline{\includegraphics[width=.5\textwidth]{figures/carbon_veg_scale.pdf}}
%\caption{Biases and mean absolute errors in carbon densities derived from cropland maps, calculated as percents relative to the reference map. Bias estimates (represented by symbols) fall within the semi-transparent bars, while mean absolute errors are contained in the solid bars. Bar colors are coded to specific cropland map, while the symbols indicate which cover type was used to calculate cropland adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values were near zero, while secondary forest values were close to forest values, and thus these are not shown for display clarity (but see Table S2).}
%\label{afoto}
%\end{figure}

>>>>

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,sort,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
%\usepackage{placeins}
%\usepackage{float}
%\restylefloat*{figure}
%\usepackage{dblfloatfix}  % http://tex.stackexchange.com/questions/235623/placing-a-figure-in-the-bottom-of-a-page-spanning-the-two-columns-of-an-ieee-doc 
%\usepackage{subfigure}
%\usepackage{wrapfig}
%\usepackage{caption}
%\setlength\belowcaptionskip{-2cm}
%\setlength{\textfloatsep}{5pt plus 1.0pt minus 2.0pt}
\setlength{\parskip}{1cm plus4mm minus3mm}

@@

\title{Landcover Data Limits Our Understanding of Earth System Processes}
%\title{Our Understanding of Global Change is Built on a Shaky Foundation}

@@

\dropcap{T}he functioning of the Earth System\footnote{give definition here} is fundamentally connected to the characteristics of landcover \cite{lambin_modelling_1997}, the physical constituents of the terrestrial surface. Human endeavors are both strongly governed by and shape landcover, whether it be felling ancient forests for timber production or burning savannas to stimulate a green flush for livestock, while landcover is a primary driver of climate and biogeochemical processes \cite{lambin_dynamics_2003}. The vastness of our modification of the Earth's surface \cite{lambin_dynamics_2003} means that socioeconomic and physical processes increasingly interact through landcover. To fully understand these processes and the nature of global change, it is therefore essential to know the nature and distribution of landcover.

This importance is understood by a growing number of social, economic, and physical scientists, who increasingly use landcover data to advance understanding in Earth System research areas ranging from food security \cite{lark_cropland_2015,wright_recent_2013, licker_mind_2010} to carbon cycling \cite{asner_high-resolution_2010, gaveau_major_2014}, biodiversity loss \cite{newbold_global_2015, luoto_predicting_2004}, and demographic shifts \cite{linard_assessing_2010}.

The validity of the knowledge resulting from such studies depends on the veracity of the landcover data upon which they are based, much as a building requires a solid foundation if it is to remain standing.  Unfortunately, the evidence so far suggests that Earth System science is being built on a shaky foundation. The reason for this is that landcover data can only practically be derived from satellite imagery, but in many regions the cover types of interest are smaller \cite[e.g. smallholder's farms][]{jain_mapping_2013} than the sensor resolution, or spectrally indistinct from neighboring covers, which translates into substantial mapping errors \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. The result is that landcover maps are generally inaccurate at finer scales and disagree substantially with one another, particularly in those parts of the world undergoing the most rapid land use changes \cite{estes_projected_2013, fritz_comparison_2010, fritz_cropland_2011}. These errors mean that we are still unable to obtain the granular, mechanistic understanding of global change processes that we need.

These problems with landcover products are known \cite{fritz_comparison_2010, fritz_cropland_2011, see_improved_2015, fritz_mapping_2015,verburg_challenges_2011}, and there are a variety of map improvement efforts underway, particularly for agriculture \cite{fritz_geo-wiki:_2012,estes_diylandcover:_2015}. What remains an open question is exactly how much the maps Earth System researchers typically use deviate from actual landcover, and how this in turn impacts our understanding of Earth System processes. Answering this question depends on having spatially comprehensive ground truth data, which are unavailable for most parts of the world, particularly over Africa and other developing regions \cite{see_improved_2015}. Our understanding of map accuracy is therefore built primarily on bottom-up tests made with a relatively small number of ground truth points (relative to the total mapped area), or from top-down "sanity checks" made in comparison to aggregated survey data. This allows us to quantify between map discrepancies  \cite[e.g.][]{fritz_comparison_2010, kaptue_tchuente_comparison_2011}, or to understand map fidelity to country-level statistics \cite[e.g.][]{fritz_comparison_2010}, but offers little direction for how to arrive at a true number.

Being unable to fully quantify the errors in landcover maps makes it difficult, if not impossible, to gauge their impact on downstream analyses. There has been some work examining how such error influences climate simulations \cite{ge_impacts_2007}, agricultural land use patterns \cite{schmit_limitations_2006}, carbon flux measurements \cite{quaife_impact_2008}, and human population estimates \cite{linard_assessing_2010}, but these either use simulated landcover errors \cite{ge_impacts_2007} or compare relevant differences in estimates between different satellite-derived landcover maps \cite{linard_assessing_2010, quaife_impact_2008}. One exception is a Belgian study \cite{schmit_limitations_2006} that used ground-collected farm parcel data to assess how landcover errors bias measurements of agricultural land use patterns, but the study extent was fairly small and the validation data were discontiguous.

Fortunately, the recent, explosive growth in public and private initiatives to develop new Earth observing capabilities, which range from small drones\footnote{e.g. 3DRobotics, DJIA} to new high resolution satellite arrays\footnote{Planet Labs, Skybox} and better mapping methods \cite{fritz_geo-wiki:_2012,estes_projected_2013,debats_generalized_????}, are finally providing the means to comprehensively interrogate the accuracy and biases in the landcover products that have become commonplace in global change research--and which are often used to make policy decisions \cite{searchinger_high_2015}.

In this study, we take advantage of these recent advances to address the urgent need to more thoroughly and precisely quantify landcover map errors and how they might impact our understanding of Earth System processes in the world's most dynamic regions.  We use a unique, high accuracy landcover map of South African crop fields to conduct a spatially comprehensive, bottom-up quantification of error in several widely used landcover maps, and how these errors can propagate through ``downstream'' studies investigating into both the physical and socioeconomic components of the Earth System. Our objective is to provide scientists and policy-makers who use landcover data with a better, more up-to-date, understanding of their appropriate uses and limitations.

\vspace{-0.5 cm}
\section{Overview of study area and analyses}
In the late 2000s, the South African government commissioned a cropland map that was made by manually interpreting and digitizing fields visible within high resolution satellite imagery \cite{fourie_better_2009}. The resulting vectorized field boundaries provide highly accurate data on field sizes and distribution for the period 2009-2011. This dataset is particularly valuable because South Africa represents nearly 6\% of sub-Saharan Africa's (SSA) area, which is a region that is poised to undergo rapid agricultural expansion \cite{searchinger_high_2015}, yet is notably lacking trustworthy maps of existing agriculture land \cite{fritz_comparison_2010}. Moreover, South Africa's large agricultural sector represents the diversity of farming systems found throughout SSA, ranging from large commercial operations to smallholder farms \cite{hardy_rainfed_2011,estes_using_2014}.

We used this dataset as a reference layer for evaluating four landcover products representative of the type commonly used in Earth Systems research. The first was South Africa's own 30 m resolution 2009 National Landcover map (SA-LC)\cite{sanbi_national_2009}, which is typical of the higher-resolution, Landsat-based maps that are typically available only for individual countries \cite[e.g.][]{fry_completion_2009}. The second and third were respectively the 300 m GlobCover 2009 \cite{arino_global_2012} and 500 m resolution 2011 MODIS Landcover products, which are widely used global-scale products \cite[e.g.][]{gross_monitoring_2013, shackelford_conservation_2015}. The fourth dataset was the new 1 km GeoWiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}, which incorporates the first three datasets and represents the current state-of-the-art in landcover mapping.  For comparison, we converted all five datasets into 1 km gridded maps of cropland density, expressed as the pixel-wise percent cover.

In our first set of comparisons, we evaluated map quality by subtracting each of the four landcover product-derived percent cropland maps (hereafter we refer to these as the ``the test maps'') from the reference map, so that we could calculate each map's bias (the mean pixel-wise error) and accuracy (the mean absolute pixel-wise error, where a lower value indicates higher accuracy). We performed this analysis for the original 1 km resolution, and for maps that were further aggregated to 5, 10, 25, 50, and 100 km, in order to evaluate how error and bias changes with scale. We also assessed how landcover pattern impacts map performance by modeling the correlation between map accuracy and cropland density.

We then used these maps to conduct four further analyses typical of Earth System science research. The first two related to physical processes, namely the calculation of vegetative carbon stocks and the simulation of evapotranspiration by a hydrological model. The second two were socio-economic in nature: the estimation of agriculture yield and production, and an agent-based model assessment of household food security.  The first and third analyses were relatively simple, in that the variable(s) of interest were mapped onto landcover using empirical relationships. The second and fourth relied on more complex numerical methods, where landcover was one of several variables needed to run each model. For the simpler analyses, we examined how results were influenced by map aggregation, while for the more complex cases, our assessments were confined to each numerical model's standard output resolution.

\vspace{-0.5 cm}
\section{Map quality}
\subsection{Bias and accuracy}

Our reference map indicated that crop fields covered 104,304 km$^2$, or nearly 10\%, of the total mapped area in the 2009-2011 time period. The test maps derived from SA-LC and GeoWiki overestimated this area by 31 and 10\%, respectively, while GlobCover and MODIS underestimated it by 18 and 23\%. Subtracting each test map from the reference maps created pixel-wise residuals, where negative and positive values respectively represent overestimates and underestimates by the test map (Fig. 1A).

\begin{figure*}[t]
%\begin{wrapfigure}{c}{1\textwidth}
\centerline{\includegraphics[width=0.95\textwidth]{figures/figure1.pdf}}
\vspace{-0.15 cm}
\caption{(A) Errors in the percent cropland estimates resulting from each of the four test maps relative to the reference map at different scale of pixel aggregation. Rows indicate the test map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas where areas under communal farmlands or permanent tree crops were removed from analysis. (B) The bias (mean error) and accuracy (mean absolute error [MAE]) of each test map at each scale of aggregation, weighted by the percentage cropland in each cell of the reference map. Bias estimates are indicated by the semi-transparent bars, accuracy (lower is more accurate) by the solid bars, with bar colors coded to specific cropland maps.}
\label{afoto1}
\end{figure*}
%\end{wrapfigure}

@@

\linenumbers
The most pronounced errors were in the MODIS and GlobCover maps, which showed large positive residuals in the center of the country where cropland is most concentrated (blue areas in Fig. 1A), and negative residuals (red areas) along the eastern and northern margins.
These patterns translated into substantial map bias (Fig. 1.B), with GlobCover and MODIS mean error (weighted by the reference cropland density) exceeding 45\% and 25\% respectively at 1 km resolution, meaning that each map tends to underestimate cropland by that amount at that resolution. This bias declined with each level of map aggregation, being reduced to nearly 15\% for GlobCover and 5\% for MODIS at 100 km. The magnitude of (density weighted) mean absolute error (MAE) was somewhat higher in all cases. The GeoWiki map, in contrast, was the least biased overall, showing just a ~7\% bias at 1 km and near 0 for all other scales of aggregation, although its accuracy (23\% MAE) was only half as good as SA-LC's at 1 km (11\% MAE), which despite its uniform overestimation bias (Fig. 1A) was the most accurate map at aggregation scales $<10$km. Above this, GeoWiki became slightly more accurate, having $<$5\% MAE at 100 km resolution. The reason GeoWiki had relatively poor accuracy at 1 km resolution was due to the heterogeneity of residuals, which traded between positive and negative residuals over short distances, thereby inflating MAE at this scale.

\subsection{Error and landcover density}
Given that landcover characteristics can influence landcover map quality \cite{see_improved_2015, estes_diylandcover:_2015, debats_generalized_????}, we set out to elucidate the relationship between cropland configuration and map error.  We first calculated the average 1 km reference cropland percentage within each of South Africa's 354 magisterial districts (the finest administrative unit, averaging 3,445 km$^2$; SI), which yielded a landscape-scaled metric of characteristic cropland density, and similarly calculated the district-wise MAE for each test map.

We then used a generalized additive model to evaluate the shape of the relationship between district MAE and cropland density. The model shows that map accuracy is typically lowest at intermediate levels of cropland density (50-60\% cover) for all but the GlobCover map (where accuracy continues to decline with cropland cover), and is highest when the landscape is dominated either by cropland or by another type (Fig. 2). In other words, accuracy is generally lowest when cropland cover is mixed evenly with other cover types. GlobCover's accuracy continued to decrease with cropland density because the dominant agricultural cover class contributing to the test map was defined as 50-70\% crops mingled with other vegetation, thus the maximum percentage was constrained by this mixture range.

\vspace{-0.75 cm}
\begin{figure}[h]
\centerline{\includegraphics[width=.5\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between map accuracy (the mean absolute error) in test maps and the actual cropland cover within agricultural landscapes (reference map pixels having $>$0.5\% cropland), here defined by the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different test maps, with the solid line indicating predicted absolute bias, and the lighter shading the standard error of the coefficients.}\label{afoto2}
\end{figure}

\section{The impact of map error on physical analyses}
\subsection{Estimating vegetative carbon stocks}
To understand the carbon cycle and climate forcing due to land use change, it is important to have accurate, high resolution maps of vegetative carbon stocks \cite[][]{searchinger_high_2015}. One widely used vegetative carbon dataset is that of \cite{ruesch_new_2008}, who mapped estimated carbon density values for different vegetation types to the classes of a global landcover product. The resulting data were intended to provide a baseline for climate policy by the Intergovernmental Panel on Climate Change (IPCC), as well as input to other land use and biogeochemical analyses \cite{ruesch_new_2008}.

%\vspace{-1 cm}
We followed this method to create vegetative carbon maps for South Africa. Since our maps represented cropland percentage, we developed several variants by assigning the carbon densities of different vegetation types (forest, secondary forest, shrubland, grassland, and sparse vegetation \cite{ruesch_new_2008}) to the non-cropland fraction of our maps. These hypothetical maps represented the range in potential carbon densities, and allowed us to investigate how carbon estimates can vary as a function of i) test map errors and ii) the properties of neighboring cover types. To assess carbon estimation error, we subtracted test map-derived carbon maps from those based on the reference map, and calculated bias and accuracy scores.

The spatial patterns of test map errors transmitted into carbon estimation errors, but the sign varied as a function of the density of carbon adjacent to croplands (SI). Where cropland was underestimated and the surrounding cover type was more carbon dense than cropland, carbon density was overestimated, but when the cover type was less dense than croplands (e.g. sparse vegetation), then carbon density was underestimated. The inverse was true where cropland was overestimated.

The magnitude of carbon errors varied as a function of the carbon density of surrounding cover, as demonstrated by the bias statistics (Fig. 3). Bias was near zero when grassland was the adjacent cover type (SI), as its carbon density is nearly the same as cropland. However, when forest was adjacent then bias was a three- to five-fold multiple of cropland map bias (Fig. 1B). At the most extreme, GlobCover's bias was -276\% at 1 km, but even SA-LC and GeoWiki had biases of 22\% and -50\%, respectively. Bias could be substantial even for the least carbon dense vegetation type (sparse), as evidenced by the 15-25\% bias at 1 km for MODIS and GlobCover under this class.  The mean bias across the different potential adjacent vegetation classes ranged between -20 for GeoWiki and -123\% for GlobCover at 1 km (with MODIS in between these), while SA-LC's average bias was 11\%.  Biases declined fairly rapidly with aggregation, with all datasets having an average (across cover types) bias magnitude of less than 10\% by 25 km of aggregation, except for GlobCover, which was -12\% at 100 km (SI).  As with cropland percentages, GeoWiki produced the least biased carbon density estimates above above 1 km resolution.

In terms of accuracy, MAE values were essentially the same as bias magnitudes, except for GeoWiki's, which were twice as large. The average MAE across vegetation classes was 47\% at 1 km, dropping to $<$10 only with 25 km of aggregation. In contrast, SA-LC's carbon estimates were twice as accurate at 1 km, and was slightly more accurate up to 25 km of aggregation were GeoWiki achieved parity.

\vspace{-0.75 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/carbon_veg_scalew.pdf}}
\caption{Biases and accuracies (mean absolute errors) of carbon densities derived from cropland maps, calculated as percents relative to the reference map. Bias estimates (represented by symbols) fall within the semi-transparent floating bars, while accuracies are contained in the solid bars. Bar colors are coded to specific cropland map, symbols indicate which cover type was used to calculate cropland-adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values were near zero, while secondary forest values were close to forest values, and thus these are not shown for display clarity (but see Table S2). MODIS and GlobCover values at 1 km exceeding the plot's Y limits are provided near their truncated tops.}
\label{afoto}
\end{figure}

\subsection{Evapotranspiration estimates}
Accurate estimation of hydrological fluxes is critical to understanding how land-atmosphere interactions impact the climate system and runoff \cite{liang_simple_1994}. Land surface hydrological models, such as the Variable Infiltration Capacity \cite{liang_simple_1994}, are used to simulate these processes, and depend on landcover maps to provide information on the characteristics of vegetation and other materials covering the surface, as these govern the rates of runoff, infiltration, and evapo-transpiration. We used the VIC model to generate 25 km estimates of monthly evapotranspiration throughout South Africa, and examined how these were impacted by error in the test maps, which were used to determine the landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components.
